\documentclass[12pt]{article}

\usepackage{mathtools}
\usepackage[a4paper,left=1.5cm,right=1.5cm]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[backend=biber, style=numeric, sorting=ynt]{biblatex}
\addbibresource{refs.bib}
\usepackage{tabularx}
\usepackage{hyperref}
\usepackage{amssymb}
\usepackage{minted}

\title{Graph drawing by stress minimization}
\author{Laslo Hauschild, Lennart Peters}
\date{\today}

\begin{document}
\maketitle

\tableofcontents
\newpage

\section{Introduction}

\section{Graph Embedding}
A graph $G=(V,E)$ is an ordered pair of vertices $V$ and a set of edges $E \subseteq \{ \{u,v \} \mid u,v \in V \}$, an edge $e$ being $e=\{u,v\}$ .
Since our ultimate goal is to have a drawn version of a given graph $G$, it is natural to limit ourselves to a finite set $V$. A finite set $V$ induces $E$ to be a finite set. First, we will define a general way  to embed graphs in vectorspaces $\mathbb{R}^n$. We will see that every graph $G$ has uncountably many possible embeddings $H$ and then optimize the look of an embedding to a given graph $G$.

\subsection{Graph Embedding}

Let $G$ be a graph. We define a map as $p : V \to A$, which maps every vertex $v \in V$ to a point $a \in A$. Let us call the image $W=Im(p) \subseteq A$ and consider $W$ as a set of vertices, then we get the following.
\begin{equation}
    \label{small-p-injective}
    p \quad \text{is injective} \iff \text{There exists exactly one bijection} \quad p=q : V \to W
\end{equation}
\linebreak Let $\Omega = \{ \{a,b \} \mid a,b \in A \} $ be the set of all unordered pairs of $W$. We define a function
\linebreak $P : E \to \Omega, \{ u,v \} \mapsto \{ p(u),p(v) \}$ and call $F=Im(P) \subseteq \Omega$. Since $p$ is injective, therefore $P$ is injective and by the same argument we get the following.
\begin{equation}
    \label{capital-p-injective} 
    P \quad \text{is injective} \iff \text{There exists exactly one bijection} \quad P=Q : E \to F
\end{equation}
\linebreak This means, an injective $p$ induces (\ref{small-p-injective}) and (\ref{capital-p-injective}). For the existence of such a $p$, we need to have
\begin{equation}
    \label{vertex-set-size}
    \quad |V| \leq |A| \text{.}
\end{equation}
\linebreak Gathering what we have, if (\ref{vertex-set-size}) is true, then there exist embeddings $p$, then $H = (p(V), P(E))$ is equivalent to $G$. To keep it short, let us say $H=p(G)$.

\subsection{Modeling stress}

We now choose $A$ to be an $\mathbb{R}$-vectorspace, s.t. we can benefit from their analytic properties. For $n \geq 1$, an injective  $ p : V \to \mathbb{R}^n$ induces an embedding of $G$ in $\mathbb{R}^n$, but we add a drawing function $d : F \to \mathbb{R}^n, \{ a,b \} \mapsto \overline{ab}$ that draws an edge as a straight-line segment. Drawing $H$ this way results in a bounded interval in $\mathbb{R}$, where straight-line segments overlap with the points, which we deem unfit for representation purposes. For $n \geq 4$, we draw in high dimensions and we will see, that $n=2,3$ suffice. When embedding a graph $G$ into $\mathbb{R} ^n$, we essentially choose a finite set of points $W$. For every graph $G$ there exist uncountably many embeddings into $\mathbb{R} ^n$. We want to gain as much order as possible. Graphs can end up looking tangled, because they are randomly chosen points by construction. A broadly used approach is to use a stress-function $S$ that assigns  the graph $H$ a real number, the stress in $H$. The local minima of $S$ are our desired representations of $G$.

\subsubsection{FR model}

Fruchterman and Reingold developed this model in 1991, following two main principles:
\begin{itemize}
\item drawing neighboring vertices near each other
\item vertices should not  be drawn too near to each other
\end{itemize}
The FR model uses the following stress-function:
\begin{equation} S(p(G)) = \frac{1}{6}\sum_{\{u,v\} \in E}{\parallel p(u)-p(v) \parallel^3}-\sum_{\{u,v \} \in V^2}{log \parallel p(u)-p(v) \parallel} \end{equation}
Observe the following:
\begin{itemize}
\item the further apart neighboring vertices get mapped the bigger the stress
\item small distances between vertices let the logarithm explode
\item the logarithm dominates at small distances and hinders the graph from collapsing, or converging into a single point
\end{itemize}
$S$ satisfies the two main principles, while being rather quick when implemented into an algorithm. The attractive force is only calculated over $F$ and the repulsive force over $W$, the authors name this as an advantage over Davidson and Harel's model, which also penalized edge crossing.

\subsubsection{ERLL model}

The Edge Repulsion LinLog model by Andreas Noack, 2004, places its focus on the graph's edges. For $v \in V$, let $deg(v)$ denote the number of edges $v$ is an endpoint of.
\begin{equation} S(p(G)) = \frac{1}{2} \sum_{\{u,v\} \in E}{\parallel p(u)-p(v) \parallel} - \sum_{\{u,v \} \in V^2}{\deg(u)\deg(v) \log\parallel p(u)-p(v) \parallel} \end{equation}
The edges cause both attraction and repulsion. Let us compare the ERLL with the FR model. Let $G$ be a graph. Now consider the unordered pairs ${u,v} \in V^2$. If $G$ is of uniform degrees, then FR and ERLL are of the same repulsion, only difference is a scalar $c=deg(v)^2$. Assume $G$ is of ununiform degrees, in FR the repulsion remains the same, but in ERLL vertices $u,v$ of high degree repel each other proportional to $deg(u)deg(v)$. The repulsive force in the FR model only depends on distance, letting "big nodes" be pulled inside the graph's embedding by the attractive force. The ERLL solves this issue, spacing "big nodes" evenly. This results in clearer embeddings of highly ununiform graphs. We still require a method to find the minima of $S$.

\subsection{Gradient Descent}

\section{Implementation}

\subsection{Plotting a graph embedding}

To plot our graph embedding, we use the Graphs library \cite{graphsjl} as well as the Plots library \cite{plotsjl}.

\begin{minted}{julia}
using Plots
using Graphs

function plot_graph(g::SimpleGraph, p::Matrix{Float64})
    plt = plot(legend = false, aspect_ratio = :equal)
    for e in edges(g)
        plot!(plt, [p[1, src(e)], p[1, dst(e)]], [p[2, src(e)], p[2, dst(e)]],
                color = :black)
    end
    scatter!(plt, p[1, :], p[2, :], markersize=5, color = :blue)
    return plt
end
\end{minted}

Note that we first plot the edges by drawing straight lines between the vertices,
and then we plot the vertices on top of the edges. This way, the vertices are not hidden behind the edges.

\subsubsection{Higher dimension plotting}

Plotting in higher dimensions is easy. Just extend the \mintinline{julia}{plot!} and \mintinline{julia}{scatter!} calls by one more argument for the third dimension:

\begin{minted}{julia}
function plot_graph_3d(g::SimpleGraph, p::Matrix{Float64})
    plt = plot(legend = false, aspect_ratio = :equal)
    for e in edges(g)
        plot!(plt, [p[1, src(e)], p[1, dst(e)]], [p[2, src(e)], p[2, dst(e)]], [p[3, src(e)], p[3, dst(e)]],
                color = :black)
    end
    scatter!(plt, p[1, :], p[2, :], p[3, :], markersize=5, color = :blue)
    return plt
end
\end{minted}

\subsection{Stress functions}

Since the two stress functions explored in the theoretical part is structured in a similar way,
we can implement them in a single function. We just need to pass the edge and vertex stress functions
as arguments. To do this, we use anonymous functions:

\begin{minted}{julia}
function stress(
    g::SimpleGraph, 
    edge_stress_f::Function, 
    vertex_stress_f::Function, 
    combine_f::Function)

    n = nv(g)
    edge_stress = 0
    for e in edges(g)
        edge_stress += edge_stress_f(src(e), dst(e))
    end
    vertex_stress = 0
    for i in 1:n
        for j in (i+1):n
            vertex_stress += vertex_stress_f(i, j)
        end
    end

    return combine_f(edge_stress, vertex_stress)
end
\end{minted}

Implementing the FR and ERLL stress functions is then straightforward:

\begin{minted}{julia}
function fr_stress(g::SimpleGraph, p::Matrix{Float64})
    D = pairwise(Euclidean(), p, dims=2)
    edge_stress_f = (i, j) -> D[i, j]^3
    vertex_stress_f = (i, j) -> log(D[i, j] + 10^(-6))
    combine_f = (x, y) -> x / 6 - y
    return stress(g, edge_stress_f, vertex_stress_f, combine_f)
end


function erll_stress(g::SimpleGraph, p::Matrix{Float64}, deg::Vector{Int})
    D = pairwise(Euclidean(), p, dims=2)
    edge_stress_f = (i, j) -> D[i, j]
    vertex_stress_f = (i, j) -> deg[i] * deg[j] * log(D[i, j] + 10^(-6))
    combine_f = (x, y) -> x / 2 - y
    return stress(g, edge_stress_f, vertex_stress_f, combine_f)
end
\end{minted}

Not that we add a small constant to the distance in the logarithm to prevent it
from exploding when two vertices are mapped to the same point.

We use the \texttt{pairwise} function from the Distances library \cite{distancesjl} to compute the distance matrix $D$.
This is way more effiecient than computing the distances in a nested loop.

\subsection{Gradient descent}

The final piece in the puzzle is to actually use the plotting and the stress functions
to find good embeddings of a given graph. To do this, we use gradient descent.
We use the Zygote library \cite{zygotejl} to compute the gradients of the stress functions.

With zygote, we can easily differentiate through our stress functions,
which are implemented in a very straightforward way. Zygote does this by using source-to-source automatic differentiation,
which means that it transforms our code into a form that can be differentiated, and then applies the chain rule to compute the gradients.
This allows us to compute the gradients of our stress functions without having to manually derive them, which can be quite
tedious and error-prone.

The principle of gradient descent is the following:

We initialize the positions of the vertices randomly in a 2D space.
We then compute the stress and its gradient at each step, and update the positions by some rule.
To check for convergence, we either stop after the maximum number of steps has been reached,
or if the stress values do not change meaningfully.
It is important that we do not check for equality of the stress values, since they are floating-point numbers and can be very close
to each other without being exactly equal. Instead, we check if the absolute difference is smaller than a small threshold (in this case, $10^{-6}$).


\subsubsection{Simple Gradient Descent}
This is the simple version of the gradient descent algorithm. One can imagine what this algorithm does by imagining a hiker
being very tired and as such, trying to find the nearest town. To do that, he always goes in the direction of steepest descent.
Crucially, he does not have any momentum.

\begin{minted}{julia}
function simple_gradient_descent(g::SimpleGraph, alpha::Float64, max_steps::Int64)
    n = nv(g) # Number of vertecies
    deg = degree(g) # The degrees of the edges
    p = rand(2, n) # Initial positions

    stresses = Float64[]

    for k in 1:max_steps
        s, (grad,) = Zygote.withgradient(q -> fr_stress(g, q), p)
        push!(stresses, s)

        if length(stresses) > 1 && abs(stresses[end] - stresses[end-1]) < 1e-6
            println("converged!")
            plot1 = plot_graph(g, p)
            plot2 = plot(stresses, label="stress")
            final = plot(plot1, plot2, layout=2)
            savefig(final, "converged.svg")
            break
        end

        p -= alpha * grad

        if k % 30 == 0
            plot1 = plot_graph(g, p)
            plot2 = plot(stresses, label="stress")
            display(plot(plot1, plot2, layout=2))
        end
    end
end
\end{minted}

\subsubsection{Nesterov's Accelerated Gradient Descent}

In this version of gradient descent, we not only go a small step in the direction of the gradient, but we also add momentum to the term.

This can be imagined by a ball rolling, which in addition to going downhill, also has some inertia, which makes it roll
faster and faster as it goes down the hill. This can help to escape local minima and find better solutions.

Friction is modelled by $\beta$, which says how much of the previous momentum we keep. Because $\beta = 0,9$, we keep 90\% of the previous momentum.

\begin{minted}{julia}
function nesterov_gradient_descent(g::SimpleGraph, alpha::Float64, beta::Float64, max_steps::Int64)
    n = nv(g) # Number of vertecies
    deg = degree(g) # The degrees of the edges
    p = rand(2, n) # Initial positions

    stresses = Float64[]

    momentum = zeros(2, n)

    for k in 0:max_steps
        s, (grad,) = Zygote.withgradient(q -> fr_stress(g, q), p + beta * momentum)
        new_momentum = beta * momentum - alpha * grad
        push!(stresses, s)

        if length(stresses) > 50 && abs(stresses[end] - stresses[end-1]) < 1e-9
            println("converged!")
            println("iteration ", k)
            plot1 = plot_graph(g, p)
            plot2 = plot(stresses, label="stress")
            final = plot(plot1, plot2, layout=2)
            savefig(final, "converged.svg")
            break
        end

        p += new_momentum
        momentum = new_momentum

        if k % 30 == 0
            plot1 = plot_graph(g, p)
            plot2 = plot(stresses, label="stress")
            display(plot(plot1, plot2, layout=2))
        end
    end
end

\end{minted}

\section{Results}

\printbibliography[heading=bibnumbered]

\end{document}